---
title: "Time Series Analysis: Final Project Proposal"
author: "Abhinav Prathipati, Sarah Hoburg"
date: "March 24, 2026"
output: 
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load necessary libraries
library(readr)
library(TSA)
library(forecast)
library(lubridate)
```
```
```{r}
brent <- read_csv("BrentOilPrices.csv", show_col_types = FALSE)

# Handle mixed date formats in the file
brent$Date <- parse_date_time(brent$Date, orders = c("d-b-y", "b d, Y"))
brent <- brent[order(brent$Date), ]

head(brent)
summary(brent$Price)

```

## Time Series Plot

```{r}
Yt <- ts(brent$Price, frequency = 252)

plot(
  brent$Date, brent$Price, type = "l",
  main = "Brent Oil Prices Over Time",
  xlab = "Date", ylab = "Price")
```


```{r}
acf(Yt, main = "Sample ACF of Brent Oil Prices")
pacf(Yt, main = "Sample PACF of Brent Oil Prices")
```

```{r}
dYt <- diff(Yt)

plot(
  dYt, type = "l",
  main = "Differenced Brent Oil Price Series",
  ylab = "Differenced Price", xlab = "Time")
```




# Introduction and Research Objectives

This section defines the scope and significance of the project.

## Problem Statement and Significance
This section describes the specific research question or issue to be investigated. It should also explain why the selected time series is important, and why its analysis is of interest.

## Objectives
The primary objective of this project is to [e.g., model the underlying stochastic process / forecast future values / identify structural breaks]. More specifically, this study aims to:
 

# Data Characterization and Exploratory Analysis

This section describes the dataset and provides an initial exploratory analysis of the time series.

## Data Source

Provide a brief description of the data source, including where the data were obtained and the sampling frequency (for example, daily, monthly, or quarterly).

## Time Series Visualization
 
